---
title: "Day 1 R review"
author: "Kara Woo"
date: "April 6, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Arithmetic

```{r arithmetic}
1 + 100
1 == 2
log(1)
```

## Creating variables

```{r vars}
x <- 1/40
log(x)
x + x
x <- 100
x
```

## Vectors

```{r}
# Create some vectors
y <- c(1, 4.5, 2.7, 9)
y
z <- 1:5
z

# Create a vector of animals
animals <- c("monkey", "rabbit", "dog")
animals

# Add 4 to each element of y
y + 4
y <- y + 4 # save y as the original y + 4
y

paste("The animal is: ", animals)
```

To inspect the type of an object:

```{r}
typeof(y)
typeof(z)
typeof(animals)
```

Predict what happens when I do this:

```{r}
new_vec <- c(5, 7.5, "hat")
typeof(new_vec)
new_vec
```

What is the value of each variable after each statement in the following:

```{r}
mass <- 47.5
age <- 122
mass <- mass * 2.3
age <- age - 20
```

Remove a variable:

```{r, error = TRUE}
y2 <- c(1, 5, 7)
rm(y2)
y2
```

## Loading data

```{r}
cats <- read.csv(file = "data/feline-data.csv")
cats

# View the class of the object (a data frame)
class(cats)

# Extract a column
cats$weight

cats$weight + 2

# Show the structure of an object
str(cats)

# Read in data without treating strings as factors
cats2 <- read.csv(file = "data/feline-data.csv", 
                  stringsAsFactors = FALSE)
str(cats2)

# Load gapminder dataset
gapminder <- read.csv(file = "data/gapminder.csv",
                      stringsAsFactors = FALSE)
```

View information about the data frame:

```{r}
# Structure of the dataset
str(gapminder)

# Number of rows, columns, rows & columns
nrow(gapminder)
ncol(gapminder)
dim(gapminder)

# Column names
colnames(gapminder)

# View first few rows
head(gapminder, n = 5)
tail(gapminder)

# Summarize data
summary(gapminder)
```

## Subsetting: Vectors

First, we'll create a named vector:

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
x
names(x) <- c("a", "b", "c", "d", "e")
x
```

Indexing:

```{r}
x[1]
x[4]
x[c(1, 3)]
x[1:4]
x[c(1, 1, 3)]
x[6]
```

Excluding:

```{r}
x[-2]
x[c(-1, -5)]
x[-c(1, 5)]
```
Subsetting by name:

```{r}
x[c("a", "c")]
```

Several ways to get b, c, and d:

```{r}
x[c(-1, -5)]
x[2:4]
x[c("b", "c", "d")]
x[c(2, 3, 4)]
```

Logical subsetting:

```{r}
x > 6
x
x[x > 6]
x[x < 7]
```

Counting how many elements of a variable meet some condition:

```{r}
length(x)
x_new <- x[x > 6]
length(x_new)
```

## Subsetting: Data Frames

Extract a column (I'll wrap this in `head` so we don't have to see all the output):

```{r}
head(gapminder$year)
```

Subset rows and columns by index

```{r}
gapminder[1:3,]
gapminder[3,]
gapminder[1:3, 1:3]
```

Subset by column names (I'll wrap this in `head` so we don't have to see all the output):

```{r}
head(gapminder[, c("lifeExp", "pop")])
```

Logical subsetting:

```{r}
head(gapminder[gapminder$year < 1960, c("year", "pop")])
```

Challenge exercise answers:

```{r}
# Extract observations collected for the year 1957 (using head() to show only the first few rows)
head(gapminder[gapminder$year == 1957,])

# Extract all columns except 1 through to 4
head(gapminder[, c(-1:-4)])
head(gapminder[, -1:-4])
head(gapminder[, -c(1:4)])

# Extract the rows where the life expectancy is longer the 80 years
head(gapminder[gapminder$lifeExp > 80, ])

# Extract the first row, and the fourth and fifth columns (lifeExp and gdpPercap).
gapminder[1, 4:5]
gapminder[1, c(4, 5)]

# Advanced: extract rows that contain information for the years 2002 and 2007
head(gapminder[gapminder$year == 2002 | gapminder$year == 2007, ])
head(gapminder[gapminder$year %in% c(2002, 2007), ])
```

## Plotting

```{r, eval = FALSE}
# Installing the package (only needs to be done once)
install.packages("ggplot2")
```

```{r}
# Load the package (needs to be done in each new R session when you want to use it)
library("ggplot2")
```

A basic scatter plot of GDP vs. life expectancy

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

\pagebreak

Life expectancy over time, points colored by continent:

```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp, color = continent)) +
  geom_point()
```

\pagebreak

Life expectancy over time, lines and points

```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp,
                             color = continent, by = country)) +
  geom_line() +
  geom_point()
```

\pagebreak

Only color the lines, not the points: 

```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp, by = country)) +
  geom_point() +
  geom_line(aes(color = continent))
```

\pagebreak

Log transformed x axis

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  scale_x_log10()
```

\pagebreak

Fit a smoothing line and increase point size:

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(size = 4) +
  scale_x_log10() +
  geom_smooth(method = "lm")
```

```{r}
# Look up help for a function
?geom_smooth
```

Facetting:

```{r, fig.asp = 1, out.width = "90%"}
# Facet grid
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~ continent)

# Facet grid
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  scale_x_log10() +
  facet_grid(year ~ continent)
```

\pagebreak

Subsetting data before plotting:

```{r}
asia <- gapminder[gapminder$continent == "Asia", ]
ggplot(data = asia, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

\pagebreak

Modifying labels and theme and saving a plot:

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  xlab("GDP per capita") +
  ylab("Life expectancy") +
  ggtitle("Figure 1") +
  theme_void() +
  ggsave("my_awesome_plot.png", width = 6, height = 4)
```

